Appendix 1: Quantifying Drought

### 2021-03-16 doc/SPEI_appendix.Rmd
port SPEI appendix to targets package style

---
title: "Appendix 1: Quantifying Drought"
author: "Eric R. Scott"
date: '2020-09-13'
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/ecology.csl" # Insert path for the bib-style
editor_options:
  markdown:
    wrap: sentence
    canonical: true
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  include = TRUE,
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "",
  dpi = 300)
library(tidyverse)
library(here)
library(conflicted)
library(SPEI)
library(tsibble)
library(lubridate)
library(glue)
library(colorspace)
library(latex2exp)
library(janitor)
library(Amelia)

library(HeliconiaDemography)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

@@

```{r embrapa-spei}
rpde <-
  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Rio Preto da Eva_1980-01-01_2016-12-31.csv")) %>% 
  select(date = Data,
         precip = Precipitacao,
         eto = Evatotranspiracao,
  )
manaus <-
  read_csv(here("analysis", "data", "raw_data", "embrapa", "Estacao_Manaus_1980-01-01_2016-12-31.csv"))%>% 
  select(date = Data,
         precip = Precipitacao,
         eto = Evatotranspiracao,
  )

embrapa_mon <-
  bind_rows("rpde" = rpde, "manaus" = manaus, .id = "station") %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  group_by(yearmonth, station) %>% 
  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
  as_tsibble(index = yearmonth, key = station)

embrapa_spei <-
  embrapa_mon %>% 
  mutate(cb = precip - eto) %>% 
  group_by(station) %>% 
  group_nest() %>% 
  mutate(st_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(st_spei, data)) %>% 
  select(-cb)

embrapa_wide <-
  embrapa_spei %>% 
  select(yearmonth, station, st_spei) %>% 
  pivot_wider(names_from = station,
              values_from = st_spei,
              names_prefix = "spei_") %>% 
  rowwise() %>% 
  mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
  ungroup()
```

@@

```{r}
# Read in data 
#Change to something like this once repo is public
# bdffp_daily <- read_csv("https://raw.githubusercontent.com/BrunaLab/Heliconia-Drought/master/data_cleaned/daily_precip.csv")
bdffp_daily <- read_csv(here("analysis", "data", "raw_data", "daily_precip.csv"))
# Prep data for imputation ------------------------------------------------

# remove accumulations
bdffp2 <-
  bdffp_daily %>% 
  filter(!flag %in% c("A", "U"))

# spread and add rows for missing dates
bdffp_wide <- 
  bdffp2 %>% 
  #complete all the dates
  as_tsibble(key = site, index = date) %>% 
  fill_gaps() %>% 
  select(date, site, precip) %>% 
  as_tibble() %>% 
  pivot_wider(names_from = site, values_from = precip) %>% 
  clean_names()

# Cluster sites by ranch to reduce missingness
bdffp_wide2 <-
  bdffp_wide %>% 
  rowwise() %>% 
  mutate(colosso_clust = mean(c(colosso, florestal, cabo_frio, gaviao),
                              na.rm = TRUE),
         km_clust = mean(c(km37, km41),
                         na.rm = TRUE)) %>% 
  mutate(across(colosso_clust:km_clust, ~ifelse(is.nan(.x), NA, .x))) %>% 
  select(-cabo_frio, -colosso, -florestal, -gaviao, -km37, -km41)

# add additional representations of time for Amelia.
full_wide <- 
  bdffp_wide2 %>% 
  mutate(
    year = year(date),
    doy = yday(date),
    yearmonth = yearmonth(date),
    .before = everything()
  )
```

```{r imputation, cache=TRUE}
# Impute missing data -----------------------------------------------------
# round(runif(1, 1, 1000))
set.seed(937)
all_cols <- colnames(select(full_wide, -year, -doy, -date))

# histograms <-
#   map(all_cols, ~{
#     ggplot(full_wide, aes(full_wide[[.x]])) + geom_histogram() + labs(title = .x)
#   })
# histograms

# variables with log-normal-ish distributions

log_cols <-
  all_cols[str_detect(all_cols, "precip") |
             all_cols %in% c("km_clust", "colosso_clust", "porto_alegre", "dimona")]

imp <- 
  amelia(
    as.data.frame(full_wide),
    p2s = 0,
    m = 10,
    ts = "doy",
    cs = "year",
    intercs = TRUE,
    polytime = 3,
    logs = log_cols,
    idvars = c("date"),
    parallel = "multicore",
    ncpus = detectCores() - 1,
    empri = .01 * nrow(full_wide) #ridge penalty because of high degree of missingness
  )
# imp$imputations[[1]]
```

```{r}
# Prep eto data for SPEI calculation
# Aggregate by month
imp_mon <-
  map(imp$imputations, ~{
    .x %>% 
      rename_with(~paste0(., ".precip"), c(dimona, porto_alegre, colosso_clust, km_clust)) %>% 
      select(date, ends_with(".precip")) %>% 
      mutate(yearmonth = tsibble::yearmonth(date)) %>% 
      group_by(yearmonth) %>% 
      summarize(across(-date, ~sum(.x, na.rm = TRUE))) %>% 
      #remove first and last month, as they aren't complete
      slice(-1, -nrow(.))
  })
# Add mean ETO from EMBRAPA data
imp_mon <-
  map(imp_mon, ~ {
    left_join(.x, embrapa_mon %>% summarize(eto = mean(eto)), by = "yearmonth")
  }) %>% 
  map(., ~{
    mutate(.x, across(ends_with(".precip"),
                      ~ .x - eto,
                      .names = "{str_remove(col, '.precip')}.cb"))
  })
# imp_mon$imp1
```

```{r}
# Calculate SPEI3 (i.e. 3-month window SPEI)

imp_spei <-
  map(imp_mon, ~{
  .x %>% 
    mutate(across(ends_with(".cb"), ~as.numeric(SPEI::spei(.x, scale = 3)$fitted),
                  .names = "{str_remove(col, '.cb')}.spei")) %>%
      select(yearmonth, ends_with(c(".precip", ".spi", ".spei")))
}) 
# imp_spei$imp1
# Combine multiply imputed SPI and SPEI results by taking mean
bdffp_spei <- 
  imp_spei %>%
  bind_rows(.id = "imp") %>% 
  #replace any infinite values with NAs
  mutate(across(ends_with(".spei"), ~ ifelse(is.infinite(.x), NA, .x))) %>% 
  group_by(yearmonth) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  #replace NaN's with NAs
  mutate(across(ends_with(".spei"), ~ ifelse(is.nan(.x), NA, .x))) %>% 
  rowwise() %>% 
  mutate(bdffp_spei = mean(c_across(ends_with(".spei")))) %>% 
  ungroup()
```

@@

```{r xa-spei}
xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
#aggregate monthly, then calculate climate water balance.
xa_mon <-
  xa %>% 
  unite(latlon, lat, lon, sep = ", ") %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  group_by(yearmonth, latlon) %>% 
  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
  as_tsibble(index = yearmonth, key = latlon)

xa_spei <-
  xa_mon %>% 
  mutate(cb = precip - eto) %>% 
  group_by(latlon) %>%
  group_nest() %>% 
  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(xa_spei, data)) %>% 
  select(-cb)
```

@@

```{r trmm-spei}
trmm <- read_csv(here("analysis", "data", "raw_data", "trmm.csv"))
trmm2 <-
  trmm %>% 
  unite(latlon, lat, long, sep = ", ") %>% 
  mutate(yearmonth = yearmonth(date), .after = date)
# fortunately xa_mon has same latlon, so can join eto

unique(xa_mon$latlon) %in% unique(trmm2$latlon)

trmm_ts <- 
  left_join(trmm2, xa_mon %>% select(yearmonth, latlon, eto), by = c("yearmonth", "latlon")) %>% 
  #I guess I only need grid cells that are also in the Xavier et al. dataset?
  filter(!is.na(eto)) %>% 
  as_tsibble(index = yearmonth, key = latlon)

trmm_spei <- 
  trmm_ts %>% 
  mutate(cb = precip - eto) %>% 
  group_by(latlon) %>% 
  group_nest() %>% 
  mutate(trmm_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(trmm_spei, data)) %>% 
  select(-cb)
```

@@

```{r global-spei}
gspei <-
  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
             delim = ";") %>% 
  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)
```

@@

```{r}
# take average across site
xa_df <-
  xa_spei %>% 
  group_by(yearmonth) %>% 
  #Summarize across grid cells
  summarize(xa_spei = mean(xa_spei))

@@

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  full_join(bdffp_spei %>% select(yearmonth, bdffp_spei)) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

@@

```{r spei-table, fig.cap=cap}
knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
```

>>>>

---
title: "Appendix 1: Quantifying Drought"
author: "Eric R. Scott"
date: '2020-09-13'
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - --lua-filter=templates/pagebreak.lua
bibliography: references.bib
csl: "templates/ecology.csl" # Insert path for the bib-style
editor_options:
  markdown:
    wrap: sentence
    canonical: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  include = TRUE,
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "",
  dpi = 300)
source(here::here("packages.R"))
```

@@

```{r embrapa-spei}
withr::with_dir(here(),{
  tar_load(embrapa_wide)
  tar_load(embrapa_mon)
})
```

@@

```{r}
withr::with_dir(here::here(), {
  tar_load(bdffp_spei)
})
```

@@

```{r xa-spei}
withr::with_dir(here::here(), {
  tar_load(xa_spei)
})
```

@@

```{r trmm-spei}
withr::with_dir(here::here(),{
  tar_load(trmm_spei)
})
```

@@

```{r global-spei}
withr::with_dir(here::here(), {
  tar_load(gspei)
})
```

@@

```{r}
# take average across site
xa_df <-
  xa_spei %>% 
  as_tibble() %>% 
  group_by(yearmonth) %>% 
  #Summarize across grid cells
  summarize(xa_spei = mean(spei))

@@

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  full_join(bdffp_spei %>% select(yearmonth, bdffp_spei)) %>% 
  pivot_longer(c(-yearmonth), names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

@@

```{r spei-table, fig.cap=cap}
# knitr::include_graphics(here("analysis", "supplementary-materials", "spei_table.png"))
table
```
